Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2023-01-11 edubots_surveys.Rnw
Reflow and some corrections in first paragraph

The key findings of this study shed light on educators’ preferences for using
messaging applications supported by their institutions. The literature on
technology adoption often highlights the importance of users' perceptions of the
technology's usefulness and ease of use for its successful implementation and
utilization. However, in the context of higher education, institutions also play
a crucial role in integrating these tools into their educational systems. This
does not only improve the uptake of these applications, but also shapes the social
and educational experiences of students. To achieve this successful integration,
institutions should ensure that these messaging applications are (in Europe)
GDPR- (General Data Protection Regulation) compliant to protect students’ data
and provide IT support to all stakeholders who use the applications. In summary,
a technology adoption strategy for instant messaging, including chatbots,
implies an institutional adoption strategy first and foremost.

Comparing these results with those obtained from student surveys in
\cite{MoraChatbots2021:anon}, we now have a clear vision of the differences
between teachers' and students' points of view and intentions when using the
messaging applications in higher education. Teachers are more likely to adopt
technologies that are supported by their institutions. This may be driven by a
desire to ensure that their universities monitor and support their efforts
to assist students during the learning process, although another reason could
simply be their familiarity with the technological services provided by their
institutions. Meanwhile, students use non-institutional messaging applications to
create informal discussion groups with their peers. It is worth noting that peer
support and collaboration are inseparable from learning
\cite{timmis2012constant} and its presence correlates positively with higher
retention rate in higher education \cite{o2014front}. Therefore, both perspectives
are complementary and play different roles in promoting the learning process.
However, there is a certain degree of incompatibility that would hinder the
process of adoption of a single messaging service that fits the need of both
collectives, or would favor the adoption of semi-informal solutions that would
rely on a popular platform (such as WhatsApp or Telegram) enhanced with
institutional agreements with their providers, as well as locally created and
supported chatbots.

@@

Based on the discussion of this study, we propose here a potential process for
introducing technology that would need to be piloted in order to determine its
value and its relationship to improved learning outcomes and higher teaching
staff satisfaction.

Initially, institutions should introduce messaging automation tools attached to
some, or several, messaging platforms that are already popular among the
university community, that would help with bureaucratic or repetitive tasks,
such as answering frequently asked questions or providing information on class
or assignment deadlines.  These chatbots will pave the way for more complex ones
that will have a greater impact on learning outcomes, such as chatbots that help
students integrate with other students in class or identify and address learning
problems in individual students or groups of them. These chatbots should also be
accompanied by analytics on student interaction and possibly some natural
language processing (in vernacular language) to help assess the general mood of
the class and how different materials or external factors affect it. These tools
would have to be attached to a learning analytics platform, which in turn,
would be part of an institutional learning management system.

The introduction of chatbots and chatbot technology will be helped by the fact
that no discernible differences are found in the survey between different
demographic groups. Even though chatbots do have some potential for
personalization or customization based on the class material, student, and
teaching staff, the institutional introduction of the technology can be done in a
general way and for all disciplines, types or degrees, teaching experience, and
gender.

Overall, the general feeling that transpires from the survey is that it is
essential for any institution to take into account stakeholders' opinions when
introducing chatbots. This is true almost across the board for any new
technology, but in the case of chatbots (and instant messaging applications)
their immediacy and the fact that they can invade what we could call the
private sphere makes this even more necessary.

One of the questions in the {\bf second survey}, which asked about the messaging
applications used by tertiary education teachers, opens a new line of inquiry
about what educators {\em perceive} as such, and how it is used.  Namely, the
responses indicated that teachers considered messaging applications not only
traditional chats or instant messaging apps such as WhatsApp or Microsoft Teams,
but also any means provided by their university for communicating with students,
such as a feature of the cmapus-wide learning management system that emails
grades to students. This suggests that teachers have a need for communication
with students, which is mostly unidirectional, and that it does not matter as
much how that need is met. However, this will need careful consideration,
including how it ties with the automation of the learning experience provided by
chatbots.

Our analysis of the survey results can inform several potential avenues for
future research. For instance, the rollout of extensive videoconferencing and
virtual teaching solutions that the COVID-19 pandemic has brought has also taught
us a series of lessons: it increases isolation and decreases the amount of
synchronous contact that happens in the fringes of the classroom (e.g.  teaching
staff offices, or before and after class). A future line of research could focus
on these issues and explore how chatbot technology could address them.

Finally, the full extent of chatbot technology is not really examined in these
surveys. They can be connected to natural language processing engines with
sentiment analysis as well as other analytics. Examining the mood of the class
such as responses to new materials, assignments, exams or external events will
help any student-centered teaching strategy, which will also help students (and
teachers) achieve their learning objectives. This line of work, however, would
require the introduction of some pilot study combined with initial opinion
assessments from students and is thus left as future work.

>>>>

The key findings of this study shed light on educators’ preferences for using
messaging applications supported by their institutions. Technology adoption
literature often focuses on users’ perceptions of the technology usefulness and
ease of use as important factors for its successful adoption and
utilization. Nevertheless, in higher education, institutions play a role in
integrating these tools into their educational systems, which can improve the uptake of
these applications and shape the social and educational experiences of 
students. To achieve this successful integration, institutions should ensure that
these messaging applications GDPR- (General Data Protection Regulation) compliant to protect students’ data and 
provide IT support to all stakeholders who use the applications.

Comparing these results with those obtained from student surveys
in \cite{MoraChatbots2021:anon}, we have better understood the differences between teachers’
and students' points of view and intentions when using the
messaging applications in higher education.  Teachers tend to adopt technologies
supported by their institutions, which could be motivated by a desire to ensure that their
universities can oversee their efforts in supporting their students during the
learning process.  Another reason could be their familiarity with the technology
provided by their institutions. Meanwhile, students use 
non-institution messaging applications to create informal discussion groups with
their peers.  It is worth noting that peer support and collaboration are
inseparable from learning \cite{timmis2012constant} and correlates positively
with higher retention in higher education \cite{o2014front}.  Therefore, both
perspectives are complementary and play different roles in promoting the
learning process.

@@

Based on the discussion of this study, we propose a potential process for introducing
technology that would need to be piloted in order to determine its value and its
relationship to improved learning outcomes and higher teacher staff satisfaction.

Initially, institutions should introduce messaging automation tools that help with bureaucratic or repetitive tasks, such as answering frequently asked questions or providing information on class or assignment deadlines. 
These chatbots will pave the way for more complex chatbots that will
have a greater impact on learning outcomes, such as chatbots that help students
integrate in class or identify and address learning problems
in individual students or groups of them. These chatbots should also be accompanied by
analytics on student interaction and possibly some natural language
processing (in vernacular language) to help assess the general mood of
the class and how different materials or external factors affect it.

The introduction of chatbots and chatbot technology will be helped by the fact
that no discernible differences are found between different groups. Even though
chatbots do have some potential for personalization or customization based on
the class material, student, and teaching staff, the general introduction of the
technology can be done in a general way and for all disciplines, teaching
experience, and gender.

Overall, the general feeling that transpires from the survey is that it is
essential for any institution to take into account shareholders' opinions
when introducing chatbots. This is true almost across the board, but in the case
of chatbots (and instant messaging applications) their immediacy and the fact
that they can enter into what we could call the private sphere makes this even
more necessary.

One of the questions in the {\bf second survey}, which asked about 
the messaging applications used by tertiary education teachers, opens a new line
of inquiry about what educators {\em perceive} as such, and how it is used.
Namely, the responses indicated that teachers perceived as messaging
application not only traditional chats or instant messaging app such
as WhatsApp or Microsoft Teams, but also the means provided by their university for
communicating with students, such as a feature of grading applications that
emails grades to students. This suggests that teachers have a need for communication with students, 
which is mostly unidirectional, and that it
does not matter as much how that need is met. However, this will need careful
consideration, including how it ties with the automation of the learning
experience provided by chatbots.

Our analysis of the survey results can inform several potential avenues for future research. 
For instance, the rollout of extensive videoconferencing and virtual
teaching solutions that the COVID pandemic has brought has also taught us a
series of lessons: it increases isolation and decreases the
amount of synchronous contact that happens in the fringes of the classroom (e.g. 
teaching staff offices, or before and after class). A future line of research could 
focus on these issues and explore how chatbot technology could address them.

Finally, the full extent of chatbot technology is not really examined in
these surveys. They can be connected to natural language processing engines with
sentiment analysis as well as other analytics. Examining the mood of the class such 
as responses to new materials, assignments, exams or external events will help
any student-centered teaching strategy, which will also help students (and
teachers) achieve their learning objectives. This line of work, however, would require
the introduction of some pilot study combined with initial opinion
assessments from students.
